\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-01 doc/publications/plos16/plos16.tex
changes to table in plos16.tex

\section{Introduction}

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater for
``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
each node in a tree thereby handling millions of features, such as in the 1000 Genomes Project data, which consists
of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools even on small
datasets where multi-threading generally performs well. 
 \marginpar{Piotr this is your 20 trees has the important
  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
  a result section.}
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

A Spark application runs on a ``driver'' node and distributes tasks to the many ``worker'' nodes, or ``executors''.

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest implementation is not
able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest
dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome 1~and~2), the time taken is
excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate
stages of building the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the
bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per
executor also increases with dimensionality due to the data types Spark ML uses, which we elaborate on in the
discussion.

@@

In the first section we explore some of the
\marginpar{Needs to be rewritten once we actually have the sections} 
properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of random
forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
respect to the dimensionality of the data, building a random forest model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI)
 and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
model.

@@

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to uninforamtive predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

Random forests  apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data sets are independently drawn. with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
 $(x_1,y_1),\ldots, (x_n,y_n)$. This si done $B$ times and  a tree is grown on each of these $B$ samples.
The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;\theta_b)=q)\right).
\end{equation*}

@@

The average prediction error for an individual tree $h(X; \theta)$ is
\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss}
\begin{equation}
PE_t = E_\theta E_{X,Y} (Y-h(X; \theta))^2.
\end{equation}
Assume that, for all $\theta$, the tree is unbiased, i.e., $EY= E_X h(X; \theta)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;\theta))$ and $(Y-h(X;\theta^\prime))$ for independent $\theta,
\theta^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

The benefits of random forest models are:
  \begin{itemize}
  \item they give an out-of-bag estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

@@

\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  However the level effect will be
apparent in the CF algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measurements at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear waht the size
of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
contitions casued by the interaction of large groups of SNPs. 
\marginpar{check this}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest.}
VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
in a ``DataFrame'', similar to DataFrames in R.

Traditionally, VariantSpark used the random forest algorithm from Spark ML. While Spark ML algorithms demonstrate
scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more
features.

@@

On the other hand, \cursedforest\ (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation
between memory and dimensionality by avoiding calculations that rely on entire feature vectors.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for every individual accross one chromosome.  
We use the phase 3 dataset which contains 2,504 individuals with approximately 81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  
For compatibility with the majority of the algorithms we demonstrate, we convert the original variant representations to double representations.  
For example, we store the absence of a variant (0\textbar0) as 0.0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1.0. and a homozygous variant (1\textbar1) as 2.0.

>>>>

\section{Introduction}
\marginpar{
we have parallelize and parallelise}

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to the many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
parallelising the split for each node in a tree thereby handling millions of features, such as in the 1000 Genomes
Project data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also
utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms
existing tools even on small datasets where multi-threading generally performs well. Harnessing the virtually unlimited
capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
variable importance selection~\cite{Szymczak2016}.

@@

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1~and~2), 
\margin[ar{something wrong here. We do not report on chromosome 1 and 2} 
the time taken is excessive due to
the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
the model (see Result~\ref{comp}).  This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in
\cursedforest\ are an inherent property of random forests. Here we provide demonstrations using synthetic data.  We then
demonstrate the scalability of \cursedforest\ in respect to the dimensionality of the data, building a random forest
model on whole-genome data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) 
\marginpar{why not just error rate}
and the error.  Finally, given the role different parameter values can play in model construction, we explore the effect tuning
these parameters can have on the prediction accuracy of the model.

@@

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to uninformative predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

Random forests  apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data sets are independently drawn, with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
 $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and  a tree is grown on each of these $B$ samples.
The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}

@@

The average prediction error for an individual tree $h(X; B)$ is
\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss}
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent $B,B^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

The benefits of random forest models are:
  \begin{itemize}
  \item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
    as an independent test set for each tree;
  \item they provide a measure of variable importance, for example by summing the number of times a variable is used to
    split a node, over all nodes and all trees. More complex measures of variable importance are also used;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

@@

\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be
apparent in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are:
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measurements at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}
GWAs data sets are an obvious application areas for this algorithm.

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
contitions casued by the interaction of large groups of SNPs. 
\marginpar{check this}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest.}
VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data in a
``DataFrame'', similar to DataFrames in R.  While Spark ML algorithms demonstrate some scalability when dealing with a large
number of samples, this scalability eventuall break down as we include more features.

@@

On the other hand, \cursedforest\  is specifically designed to handle wide ``cursed'' data. It avoids the relation
between memory and dimensionality by avoiding calculations that rely on entire feature vectors.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous variant (1\textbar0 or 0\textbar1)as 1,  and a
homozygous variant (1\textbar1) as 2.
